Comparison of FDR methods on 16S microbiome data with SummarizedBenchmark

### 2017-12-18 datasets/microbiome/microbiome-SummarizedBenchmark.Rmd
Remove ash from being calculated

```{r, schubert-readdata}
otu_path = paste0(datdir, "cdi_schubert_results/RDP/cdi_schubert.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(datdir, "cdi_schubert_results/cdi_schubert.metadata.txt")

@@

```{r, schubert-cleandata}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) > n_reads])
  }

@@

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 1e-6
abun_otu <- abun_otu + zeroabun

@@

```{r, schubert-pvals}
resfile <- paste0(resdir, "schubert_results_",
                  nrow(abun_otu), "_OTUs.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("CDI", "nonCDI")
  ctrl_idx <- meta$DiseaseState %in% c("H")

@@

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  save(res, file=resfile)

@@

```{r, schubert-addphylo}
res <- res %>% separate(otu, 
                        c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
                        sep=";", remove = FALSE)

@@

```{r, schubert-ubi, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="ubiquity", maxy=20, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="ubiquity")
```

@@

```{r, schubert-abun, fig.width=10, fig.height=3.5}
strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=8, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="mean_abun_present")
```

@@

```{r, schubert-phylo, fig.width=10, fig.height=6}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

```{r, schubert-labelcovariate}
res <- res %>% mutate(ind_covariate = ubiquity)
```

@@

```{r, schubert-sb, results="hide", message=FALSE, eval=FALSE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, schubert-metrics, eval=FALSE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, schubert-plot, results="hide", width=15, height=15, eval=FALSE}
# don't plot scott theoretical for now
rmv.scott <- which(grepl("scott", colnames(sb)))

# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb, as_fraction=FALSE, supplementary=FALSE)

rejection_scatter_bins(sb, covariate="ind_covariate", bins=4, supplementary=FALSE)

# upset plot (only plot IHW for the plotted alpha)
rmv.ihw <- which(grepl("ihw", colnames(sb)) &
                              colData(sb)$param.alpha != 0.05)
rmv.cols <- c(rmv.ihw, rmv.scott)

plotMethodsOverlap(sb[,-rmv.cols], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.cols),
                   order.by="freq", decreasing=TRUE)

# upset plot excluding "unadjusted"
plotMethodsOverlap(sb[,-c(1,rmv.cols)], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.cols)-1,
                   order.by="freq", decreasing=TRUE)
```

@@

```{r, eval=FALSE}
assays(sb)[["qvalue"]][ , 15] %>% max()
sum(is.na(assays(sb)[["qvalue"]][ , 22]))
sum(is.na(assays(sb)[["qvalue"]][ , 20]))
```

Rather than de-bug this dataset (which might be weird), let's try instead to work with the obesity dataset and see if I get the same problems.

@@

```{r, schubert-cleanup}
rm(res)
rm(bd)
rm(sb)
rm(pf)
```

@@

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  save(res, file=resfile)

@@

@@

```{r, goodrich-sb, results="hide", message=FALSE, eval=FALSE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, goodrich-metrics, eval=FALSE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, goodrich-plot, results="hide", width=15, height=15, eval=FALSE}

# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb, as_fraction=FALSE, supplementary=FALSE)

rejection_scatter_bins(sb, covariate="ind_covariate", bins=4, supplementary=FALSE)

# upset plot (only plot IHW for the plotted alpha)
rmv.ihw <- which(grepl("ihw", colnames(sb)) &
                              colData(sb)$param.alpha != 0.05)

plotMethodsOverlap(sb[,-rmv.ihw], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.ihw),
                   order.by="freq", decreasing=TRUE)

# upset plot excluding "unadjusted"
plotMethodsOverlap(sb[,-c(1,rmv.ihw)], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.ihw)-1,
                   order.by="freq", decreasing=TRUE)
```

Hm, looks like all of the methods run without giving NaN's with nmids = 150 or lower...
But the results still aren't right.

@@

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 1e-7
abun_otu <- abun_otu + zeroabun

@@

@@

```{r, baxter-sb, results="hide", message=FALSE, eval=FALSE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, baxter-metrics, eval=FALSE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, baxter-plot, results="hide", width=15, height=15, eval=FALSE}
# don't plot scott theoretical for now
rmv.scott <- which(grepl("scott", colnames(sb)))

# plot nrejects by method overall and stratified by covariate
rejections_scatter(sb, as_fraction=FALSE, supplementary=FALSE)

rejection_scatter_bins(sb, covariate="ind_covariate", bins=4, supplementary=FALSE)

# upset plot (only plot IHW for the plotted alpha)
rmv.ihw <- which(grepl("ihw", colnames(sb)) &
                              colData(sb)$param.alpha != 0.05)
rmv.cols <- c(rmv.ihw, rmv.scott)

plotMethodsOverlap(sb[,-rmv.cols], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.cols),
                   order.by="freq", decreasing=TRUE)

# upset plot excluding "unadjusted"
plotMethodsOverlap(sb[,-c(1,rmv.cols)], 
                   alpha=0.05, nsets=ncol(sb)-length(rmv.cols)-1,
                   order.by="freq", decreasing=TRUE)
```

>>>>

```{r, schubert-readdata, eval=TRUE}
otu_path = paste0(datdir, "cdi_schubert_results/RDP/cdi_schubert.otu_table.100.denovo.rdp_assigned")
meta_path = paste0(datdir, "cdi_schubert_results/cdi_schubert.metadata.txt")

@@

```{r, schubert-cleandata, eval=TRUE}
remove_shallow_smpls <- function(df, n_reads) {
  # Removes samples with fewer than n_reads from dataframe df.
  # df has OTUs in rows and samples in columns
  return(df[, colSums(df) > n_reads])
  }

@@

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 0
abun_otu <- abun_otu + zeroabun

@@

```{r, schubert-pvals, eval=TRUE}
resfile <- paste0(resdir, "schubert_results_",
                  nrow(abun_otu), "_OTUs.RData")
if (!file.exists(resfile)){
  # Get case and control indices
  case_idx <- meta$DiseaseState %in% c("CDI", "nonCDI")
  ctrl_idx <- meta$DiseaseState %in% c("H")

@@

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, wilcox_teststat = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

  save(res, file=resfile)

@@

```{r, schubert-addphylo, eval=TRUE}
res <- res %>% separate(otu, 
                        c("kingdom", "phylum", "class", "order", "family", "genus", "species", "denovo"), 
                        sep=";", remove = FALSE)

@@

```{r, schubert-ubi, fig.width=10, fig.height=3.5, eval=TRUE}
strat_hist(res, pvalue="pval", covariate="ubiquity", maxy=20, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="ubiquity")
```

@@

```{r, schubert-abun, fig.width=10, fig.height=3.5, eval=TRUE}
strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=8, binwidth=0.05)
rank_scatter(res, pvalue="pval", covariate="mean_abun_present")
```

@@

```{r, schubert-phylo, fig.width=10, fig.height=6, eval=TRUE}
#strat_hist(res, pvalue="pval", covariate="mean_abun_present", maxy=20)
#rank_scatter(res, pvalue="pval", covariate="ubiquity")
ggplot(res, aes(x=pval)) + geom_histogram() + facet_wrap(~phylum, scales = "free")
```

@@

```{r, schubert-labelcovariate, eval=TRUE}
res <- res %>% mutate(ind_covariate = ubiquity)
```

Also (hacky), remove test_statistic so that ash doesn't run.

```{r, schubert-rm-ash, eval=TRUE}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

@@

```{r, schubert-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, schubert-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, schubert-plot, width=15, height=15, eval=TRUE}

# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )

@@

```{r, eval=FALSE}
assays(sb)[["qvalue"]]["ashs"] %>% max()
sum(is.na(assays(sb)[["qvalue"]]["lfdr"]))
sum(is.na(assays(sb)[["qvalue"]]["scott-empirical"]))
sum(is.na(assays(sb)[["qvalue"]]["scott-theoretical"]))
```

Plotting methods are giving errors for some reason. Let's try to use Alejandro's code instead.

@@

```{r, schubert-cleanup, eval=FALSE}
rm(res)
rm(bd)
rm(sb)
rm(pf)
```

@@

  res <- data.frame(otu = rownames(abun_otu), 
                    pval = pvals, test_statistic = teststats,
                    SE = ses, effect_size = effs, 
                    mean_abun = mean_abuns, mean_abun_present = mean_abuns_present,
                    ubiquity = ubis)
  res <- res %>%
    mutate(test_statistic = qnorm(exp(log(pval)-log(2)), lower.tail=FALSE) * sign(effect_size))

  save(res, file=resfile)

@@

Also (hacky), remove test_statistic so that ash doesn't run.

```{r, goodrich-rm-ash}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

@@

```{r, goodrich-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, goodrich-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, goodrich-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
#plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE )
```

There is no upsetR plot for OB Goodrich for now because only lfdr is returning any rejections.

@@

# Add pseudo counts
#minabun <- min(abun_otu[abun_otu > 0]) # Use this to figure out what pseudo-count value to add
zeroabun <- 0
abun_otu <- abun_otu + zeroabun

@@

Also (hacky), remove test_statistic so that ash doesn't run.

```{r, baxter-rm-ash}
res <- res %>% 
  mutate(test_statistic = NA) %>%
  mutate(effect_size = NA)
```

@@

```{r, baxter-sb, results="hide", message=FALSE, eval=TRUE}
for (nmids in c(5, 50, 150)){
  print(nmids)
  bd <- initializeBenchDesign(nmids=nmids)

@@

```{r, baxter-metrics, eval=TRUE}
# rename assay to qvalue
assayNames(sb) <- "qvalue"
sb <- addDefaultMetrics(sb)
```

@@

```{r, baxter-plot, width=15, height=15, eval=TRUE}
# plot nrejects by method overall and stratified by covariate
rejections_scatter( sb, as_fraction=FALSE, supplementary=FALSE)
rejection_scatter_bins(sb, covariate="ind_covariate", supplementary=FALSE)
#plotFDRMethodsOverlap(sb, alpha=0.1, supplementary=FALSE, order.by="freq", nsets=100 )
```
